Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-06-14 SIR_manuscript.tex
adding Faddy's work, changed the coeff from ABC to abc_4

\item \label{p1:eta} Increasing isolation efficacy for waiting ($\theta\_w$) and confirmed-positive ($\theta\_c$) individuals always decreases \Rnum (Eqs.~\ref{eq:del4_theta},~\ref{eq:d_del4_thetac},~\ref{eq:d_del4_thetaw});
\item \label{p1:rho} Higher testing intensity $\rho$ decreases $\Rnum$ if
%% \begin{enumerate} \item
testing is random (all $w_X$ equal) or testing intensity ($\rho$) is small (Eq.~\ref{eq:dd3dr}).
%% \end{enumerate}
\item \label{p1:omega} Increasing the rate of test return ($\omega$) always decreases \Rnum if waiting individuals do not isolate ($\theta_w=0$) (Eq.~\ref{eq:dlindo}).
\item \label{p1:w} Increasing testing focus, i.e. changing the testing weights from random ($w_S=w_I$) toward targeted  ($w_S<w_I$), always decreases \Rnum (Eq.~\ref{eq:d_del_wis}).
\end{enumerate}

\newpage
\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf Effectiveness of testing and isolation in reducing $\Rnum$ at low \percap testing intensity.}
Numerical evaluation of the effectiveness of control ($\Delta$: eq.~\ref{eq:del4}), over a range of testing and isolation parameters. Parameter values (Table~\ref{tab:params}):
$\beta=\betaparam$ 1/day, $1/\gamma= \invgammaparam$ days (baseline $\Rnum=\Rnumparam$, $r=\rparam$); $\omega \in [1/12,2]$ 1/day;  $\rho \in [0,0.013]$ 1/day per capita; $\theta\_w$ and $\theta\_c$ vary between 0 (no effect of isolation) and 1 (complete elimination of transmission); $p_S=0$, $p_I=1$ and $p_R=0.5$. Only parameter sets where $\theta_c \geq \theta_w$ (confirmed-positive individuals isolate more effectively than waiting individuals) are shown; the alternative case, $\theta_w > \theta_c$, is unrealistic. Contours of $\Delta$ are plotted for (a) random testing ($w_S=w_I=w_R=1$) and (b) targeted testing ($w_S=0.3$; $w_I=w_R=1$). 
}
\label{pan}
\end{figure}

@@

\begin{itemize}
\item We would generally expect increasing testing delays to increase \Rnum, thus decreasing effectiveness of control $\Delta$. This is in fact what happens when waiting individuals do not isolate ($\theta_w =0$, top row of \fref{pan}) --- as we move to the right within each plot in this row, $\Delta$ decreases.
However, when waiting individuals isolate ($\theta_w>0$), we more often see the opposite effect: longer testing delays lead to a greater control effect $\Delta$. This is because people waiting for negative tests are assumed to continue to isolate; this applies both to susceptibles and to people who became infected while waiting for negative test results. This effect outweighs the effect of confirmed individuals isolating except when this isolation parameter $\theta_c$ is substantially bigger than $\theta_w$. This result depends on the idea that, all else equal, people who have to wait longer for test results isolate at the same level (but for a longer time) as they would if the wait were shorter.
\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). This relationship can be reversed, however, at very high testing intensities. This can only occur when testing is targeted, and $\theta_w$ is relatively small (\fref{pan2}(b), right three panels of top row). It is theoretically possible for increasing testing intensity to \emph{increase} \Rnum because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; if these people become infected while waiting, they will need to wait for their negative test result before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.

@@

Over a broad range of parameter space (for random testing, approximately $\theta_w \geq 0.25$; for targeted testing, $\theta_w \geq 0.25$ and either $\theta_c \ge 0.5$ or $1/\omega > 5$), decreasing $\omega$ --- i.e., decreasing the rate at which tested people find out about their infection status --- decreases $\Rnum$. This result is counterintuitive; public health agencies have invested a lot of effort in increasing this rate. Dynamically, this effect occurs because speeding up test returns shortens the isolation period of uninfected individuals. For infected people it only shortens the time to progression to the isolation level of the confirmed-positive compartment. Slowing test returns increases $\Rnum$ only if the proportion of infectives in the tested population (test positivity), and the magnitude of increase in isolation effectiveness from the waiting to the confirmed-positive compartment, are large enough to outweigh the decrease in isolation of uninfected waiting individuals.

While slowing test returns does decrease $\Rnum$ over a broad range of parameter space in our model, there are several real-world processes missing from our model that make it implausible that slowing test returns would actually be an effective public health measure. First, our model is missing the components that would allow us to model the primary benefit of rapid testing, i.e. detecting and containing outbreaks while they are still in progress. This process could be modeled phenomenologically by making the testing focus an increasing function of the speed of test returns (i.e. $\partial w_I(\omega)/\partial \omega > 0$), because rapid testing allows more tests to be concentrated on infected individuals. Second, in the real world individuals become less likely to maintain isolation if they are required to do so for longer; phenomenologically, we could allow  effectiveness of isolation while waiting for test returns an increasing function of test return speed ($\partial \theta_W(\omega)/\partial \omega > 0$), or we could introduce a separate ``waiting, but no longer isolating'' compartment that individuals entered from the ``waiting, isolated'' compartment at a specified rate. Third and finally, if one wants to decrease the overall transmission rate of the population there are more effective methods than keeping tested people in limbo: masking, ventilation, distancing measures, retail and event closures, stay-at-home orders, \ldots

The other counterintuitive result from our analysis is that, for sufficiently high testing intensity $\rho$, increasing testing intensity can actually increase $\Rnum$ (e.g. \fref{pan2}(b), upper right panel [$\theta_c=1$, $\theta_w=0$]). This phenomenon occurs mainly because of a peculiar feature of the DFE in our model: we assume that testing is ongoing before the beginning of the epidemic, so that the starting point of the epidemic invasion is an equilibrium distribution of susceptibles between the $S\_n$ (untested) and $S\_u$ (waiting) compartments: in particular, $S\_u^* = \rho/(\rho + \omega)$. (In reality, testing intensity itself is a dynamic variable that will increase as the epidemic proceeds.) If isolation in the waiting compartment is low ($\theta_w \ll 1$), then waiting individuals can easily be infected once the epidemic starts. Once infected they can still infect others while in the waiting compartment, and cannot be tested again before they have returned to the untested compartment. Thus, under certain extreme conditions (low $\theta_w$, high $\theta_c$, low $\omega$, high $\rho$), $\Rnum$ becomes an increasing function of testing intensity. \bmb{This phenomenon occurs primarily (only??) for targeted testing, because \ldots CAN WE FIGURE OUT WHY? Targeting should be irrelevant to determining the DFE (because everyone is $S$ at that point), so what is the dynamical explanation? Is there a second order effect somewhere, e.g. testing is more effective overall (darker colours in Fig 3(b) upper-right panel than in Fig 3(a) upper panel), so we more quickly reach the point where the effect of increasing $S\_n^*$ outweighs the beneficial effects of testing?} However, this phenomenon is even more unrealistic than the possibility that slowing test returns increases $\Rnum$. It depends both on the assumption that regular testing is occurring before the epidemic starts, and on levels of testing that are unrealistically high (at least in large, general-population settings), i.e. at least 10\% of the population being tested per day.

@@

% %%%%%%%
\subsection{rate of returning tests} \label{app:omega}
{\bf The third goal} is to explore how changes in the rate of test return $\omega$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder\Rnum{\omega}$, which specifically depends on $\pder\Delta{\omega}$. We use
the linearization of $\Delta$ around $\rho=0$ to show that there a non-monotonic relationship between $\Rnum$  and $\omega$. The linear term in the Taylor expansion of $\Delta$ at $\rho=0$ is
\begin{equation}
\label{eq:lin}
\Delta = \frac{\rho}{\omega \gamma(\omega+\gamma)} \Big(
\frac{w_I}{w_S} \omega^2 p_I \theta\_c+(\frac{w_I}{w_S}+1) \gamma\omega\theta\_w +\gamma^2 \theta\_w(2-\theta\_w)
\Big). 
\end{equation}
This results in
\begin{equation}
\label{eq:dlindo}
\pder\Delta{\omega} = \frac{\rho}{\omega^2 (\omega+\gamma)^2} \Big(
(p_I \frac{w_I}{w_S}\theta\_c -(1+\frac{w_I}{w_S})\theta\_w)\omega^2+2\theta_w \gamma(\theta_w-2) \omega+\theta_w \gamma^2 (\theta_w-2)
\Big),
\end{equation}
around $\rho=0$.
% \ali{I stoped here!}

>>>>

\item \label{p1:eta} Increasing isolation efficacy for waiting ($\theta\_w$) and confirmed-positive ($\theta\_c$) individuals always decreases \Rnum (Eqs.~\ref{eq:del4_theta},~\ref{eq:d_del4_thetac},~\ref{eq:d_del4_thetaw});
\item \label{p1:rho} Higher testing intensity $\rho$ decreases $\Rnum$ if
%% \begin{enumerate} \item
testing is random (all $w_X$ equal) or testing intensity ($\rho$) is small (Eq.~\ref{eq:dd3dr}).
%% \end{enumerate}
\item \label{p1:omega} Increasing the rate of test return ($\omega$) always decreases \Rnum if waiting individuals do not isolate ($\theta\_w=0$) (Eq.~\ref{eq:dlindo}).
\item \label{p1:w} Increasing testing focus, i.e. changing the testing weights from random ($w_S=w_I$) toward targeted  ($w_S<w_I$), always decreases \Rnum (Eq.~\ref{eq:d_del_wis}).
\end{enumerate}

\newpage
\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf Effectiveness of testing and isolation in reducing $\Rnum$ at low \percap testing intensity.}
Numerical evaluation of the effectiveness of control ($\Delta$: eq.~\ref{eq:del4}), over a range of testing and isolation parameters. Parameter values (Table~\ref{tab:params}):
$\beta=\betaparam$ 1/day, $1/\gamma= \invgammaparam$ days (baseline $\Rnum=\Rnumparam$, $r=\rparam$); $\omega \in [1/12,2]$ 1/day;  $\rho \in [0,0.013]$ 1/day per capita; $\theta\_w$ and $\theta\_c$ vary between 0 (no effect of isolation) and 1 (complete elimination of transmission); $p_S=0$, $p_I=1$ and $p_R=0.5$. Only parameter sets where $\theta\_c \geq \theta\_w$ (confirmed-positive individuals isolate more effectively than waiting individuals) are shown; the alternative case, $\theta\_w > \theta\_c$, is unrealistic. Contours of $\Delta$ are plotted for (a) random testing ($w_S=w_I=w_R=1$) and (b) targeted testing ($w_S=0.3$; $w_I=w_R=1$). 
}
\label{pan}
\end{figure}

@@

\begin{itemize}
\item We would generally expect increasing testing delays to increase \Rnum, thus decreasing effectiveness of control $\Delta$. This is in fact what happens when waiting individuals do not isolate ($\theta\_w =0$, top row of \fref{pan}) --- as we move to the right within each plot in this row, $\Delta$ decreases.
However, when waiting individuals isolate ($\theta\_w>0$), we more often see the opposite effect: longer testing delays lead to a greater control effect $\Delta$. This is because people waiting for negative tests are assumed to continue to isolate; this applies both to susceptibles and to people who became infected while waiting for negative test results. This effect outweighs the effect of confirmed individuals isolating except when this isolation parameter $\theta\_c$ is substantially bigger than $\theta\_w$. This result depends on the idea that, all else equal, people who have to wait longer for test results isolate at the same level (but for a longer time) as they would if the wait were shorter.
\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). This relationship can be reversed, however, at very high testing intensities. This can only occur when testing is targeted, and $\theta\_w$ is relatively small (\fref{pan2}(b), right three panels of top row). It is theoretically possible for increasing testing intensity to \emph{increase} \Rnum because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; if these people become infected while waiting, they will need to wait for their negative test result before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.

@@

Over a broad range of parameter space (for random testing, approximately $\theta\_w \geq 0.25$; for targeted testing, $\theta\_w \geq 0.25$ and either $\theta\_c \ge 0.5$ or $1/\omega > 5$), decreasing $\omega$ --- i.e., decreasing the rate at which tested people find out about their infection status --- decreases $\Rnum$. This result is counterintuitive; public health agencies have invested a lot of effort in increasing this rate. Dynamically, this effect occurs because speeding up test returns shortens the isolation period of uninfected individuals. For infected people it only shortens the time to progression to the isolation level of the confirmed-positive compartment. Slowing test returns increases $\Rnum$ only if the proportion of infectives in the tested population (test positivity), and the magnitude of increase in isolation effectiveness from the waiting to the confirmed-positive compartment, are large enough to outweigh the decrease in isolation of uninfected waiting individuals.

While slowing test returns does decrease $\Rnum$ over a broad range of parameter space in our model, there are several real-world processes missing from our model that make it implausible that slowing test returns would actually be an effective public health measure. First, our model is missing the components that would allow us to model the primary benefit of rapid testing, i.e. detecting and containing outbreaks while they are still in progress. This process could be modeled phenomenologically by making the testing focus an increasing function of the speed of test returns (i.e. $\partial w_I(\omega)/\partial \omega > 0$), because rapid testing allows more tests to be concentrated on infected individuals. Second, in the real world individuals become less likely to maintain isolation if they are required to do so for longer; phenomenologically, we could allow  effectiveness of isolation while waiting for test returns an increasing function of test return speed ($\partial \theta\_w(\omega)/\partial \omega > 0$), or we could introduce a separate ``waiting, but no longer isolating'' compartment that individuals entered from the ``waiting, isolated'' compartment at a specified rate. Third and finally, if one wants to decrease the overall transmission rate of the population there are more effective methods than keeping tested people in limbo: masking, ventilation, distancing measures, retail and event closures, stay-at-home orders, \ldots

The other counterintuitive result from our analysis is that, for sufficiently high testing intensity $\rho$, increasing testing intensity can actually increase $\Rnum$ (e.g. \fref{pan2}(b), upper right panel [$\theta\_c=1$, $\theta\_w=0$]). This phenomenon occurs mainly because of a peculiar feature of the DFE in our model: we assume that testing is ongoing before the beginning of the epidemic, so that the starting point of the epidemic invasion is an equilibrium distribution of susceptibles between the $S\_n$ (untested) and $S\_u$ (waiting) compartments: in particular, $S\_u^* = \rho/(\rho + \omega)$. (In reality, testing intensity itself is a dynamic variable that will increase as the epidemic proceeds.) If isolation in the waiting compartment is low ($\theta\_w \ll 1$), then waiting individuals can easily be infected once the epidemic starts. Once infected they can still infect others while in the waiting compartment, and cannot be tested again before they have returned to the untested compartment. Thus, under certain extreme conditions (low $\theta\_w$, high $\theta\_c$, low $\omega$, high $\rho$), $\Rnum$ becomes an increasing function of testing intensity. \bmb{This phenomenon occurs primarily (only??) for targeted testing, because \ldots CAN WE FIGURE OUT WHY? Targeting should be irrelevant to determining the DFE (because everyone is $S$ at that point), so what is the dynamical explanation? Is there a second order effect somewhere, e.g. testing is more effective overall (darker colours in Fig 3(b) upper-right panel than in Fig 3(a) upper panel), so we more quickly reach the point where the effect of increasing $S\_n^*$ outweighs the beneficial effects of testing?} However, this phenomenon is even more unrealistic than the possibility that slowing test returns increases $\Rnum$. It depends both on the assumption that regular testing is occurring before the epidemic starts, and on levels of testing that are unrealistically high (at least in large, general-population settings), i.e. at least 10\% of the population being tested per day.

@@

% %%%%%%%
\subsection{Rate of returning tests} \label{app:omega}
{\bf The third goal} is to explore how changes in the rate of test return $\omega$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder\Rnum{\omega}$, which specifically depends on $\pder\Delta{\omega}$. We use
the linearization of $\Delta$ when $\rho \ll 1$ to show that there a non-monotonic relationship between $\Rnum$  and $\omega$. The linear term in the Taylor expansion of $\Delta$ when $\rho \ll 1$ is
\begin{equation}
\label{eq:lin}
\Delta = \frac{\rho}{\omega \gamma(\omega+\gamma)} \Big(
\frac{w_I}{w_S} \omega^2 p_I \theta\_c+(\frac{w_I}{w_S}+1) \gamma\omega\theta\_w +\gamma^2 \theta\_w(2-\theta\_w)
\Big). 
\end{equation}
This results in
\begin{equation}
\label{eq:dlindo}
\pder\Delta{\omega} = \frac{\rho}{\omega^2 (\omega+\gamma)^2} \Big(
(p_I \frac{w_I}{w_S}\theta\_c -(1+\frac{w_I}{w_S})\theta\_w)\omega^2-2\theta\_w \gamma(2-\theta\_w) \omega+\theta\_w \gamma^2 (\theta\_w-2)
\Big).
\end{equation}
% ALi & Faddy's stuff
The latter expression has two roots
\begin{align}
\label{eq:omega_roots}
\omega^*_{+} &= \frac{\gamma\Big(-c_4 + \sqrt{c_4^2 + (a_4-b_4)c_4}\Big)}{b_4-a_4} \\
\omega^*_{-} &= \frac{\gamma\Big(-c_4 - \sqrt{c_4^2 + (a_4-b_4)c_4}\Big)}{b_4-a_4},
\end{align}
where $a_4 = p_I\frac{w_I}{w_S}\theta\_c$, $b_4 = (1+\frac{w_I}{w_S})\theta\_w$, $c_4 = \theta\_w(2-\theta\_w)$. This enables us to describe the behaviour of $\Rnum$ with respect to $\omega$ in the foolowing two cases.
\begin{itemize}
\item \textbf{Case I:} If $b_4 \geq a_4$, $\pder\Delta{\omega} < 0$, so $\Rnum$ is always increasing with respect to $\omega$ (i.e., it is always harmful to return tests more rapidly). 
\item \textbf{Case II:} If $b_4 < a_4$, $\Rnum$ will be decreasing with respect to $\omega$ (i.e., returning tests more rapidly is beneficial) only when $\omega > \omega^*_-.$
\end{itemize}
Note that $b_4 \geq a_4$ is characterized by
\begin{equation}\label{case_1_ineq}
\Big(\frac{w_I}{w_S}+1\Big)\theta\_w \geq \frac{w_I}{w_S}p_I\theta\_c \Longleftrightarrow \Big(\frac{w_S}{w_I}+1\Big)\frac{1}{p_I} \geq \frac{\theta\_c}{\theta\_w}.
\end{equation}
We begin with a \textbf{proof of Case I}. Suppose that $b_4 \geq a_4$. If the roots of $\pder\Delta{\omega}$ are not complex, then we must have $c_4 \geq b_4-a_4$. Note that $\omega^*_{-}$ must be negative since the numerator is clearly negative but the denominator is positive. Next, note that since $c_4 \geq b_4-a_4$, the numerator of $\omega^*_{+}$ must also be negative, so $\omega^*_{+}$ is negative. Thus, in this case, $\pder\Rnum{\omega}$ does not change sign on $(0,\infty)$. Checking the sign of \ref{eq:dlindo} for arbitrarily large $\omega$ shows that it is negative (since the $(a_4-b_4)\omega^2$ term dominates and is negative). So $\Rnum$ is increasing with respect to $\omega$ on all of $(0,\infty)$.

We now turn our attention to a \textbf{proof of Case II}. Suppose that $b_4 < a_4$. It follows that $\omega^*_+$ and $\omega^*_-$ are real since $c_4 > 0 > b_4-a_4$. Next, note that $\omega^*_-$ is positive since both the numerator and denominator are negative. On the other hand, $\omega^*_+$ is negative since the denominator is negative but the numerator is positive (because $c_4 > b_4-a_4$). Thus, our task is to understand the sign of $\pder\Delta{\omega}$ around the root $\omega^*_-$. Checking the sign of \ref{eq:dlindo} for arbitrarily large $\omega$ shows that it is positive (since the $(a_4-b_4)\omega^2$ term dominates and is positive). Likewise, checking the sign for values of $\omega$ close to $0$ shows that it is negative. Thus, $\Rnum$ is increasing with respect to $\omega$ when $\omega < \omega^*_-$, and is decreasing $\omega > \omega^*_-$.

Having presented the formal analysis, we now concern ourselves with its \textbf{biological interpretations.} We begin by interpreting \ref{case_1_ineq}, under which returning tests more rapidly is \emph{always} harmful. Notice that the ratio $\frac{\theta\_c}{\theta\_w}$ is simply a measure of how much more strongly individuals self-isolate when they test positive compared to when waiting for tests. Since the rate of test return directly influences the rate at which individuals change from a waiting state to a confirmed-positive state, it is intuitive that $\frac{\theta\_c}{\theta\_w}$ would appear in \ref{case_1_ineq}. Next, note that the left-hand side increases when test sensitivity decreases and when targeting of positive individuals is poor. This is consistent with our intuition: a false negative that is returned more rapidly may encourage an infectious individual to not carefully self-isolate, thus increasing transmission. Likewise, if individuals tested are mainly susceptible (rather than infectious), then returning tests more slowly would encourage them to self-isolate for longer while awaiting test results. Having understood the role of each of the parameters in \ref{case_1_ineq}, a ``holistic" interpretation of this inequality is that returning tests more slowly is helpful when the benefit of extended self-isolation (due to awaiting test results) outweighs the benefit of identifying positive cases.

Now we interpret \textbf{Case II}.  In this case, $\Rnum$ will have a global maximum with respect to $\omega$ at $\omega^*_-$. Note that our model assumption that $\rho < \omega$ plays an important role here: if $\omega^*_- < \rho$, then $\Rnum$ will be always decreasing with respect to $\omega$. On the other hand, if $\rho < \omega^*_-$, then $\Rnum$ will be increasing with respect to $\omega$ on $(\rho, \omega^*_-)$ and decreasing beyond that.
